Nonparametric Clustering Based on Energy Statistics

### 2017-05-16 qcqp_clustering/draft/energy_clustering.tex
draft

\newcommand\Energy{\mathcal{E}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\Zt{\widetilde{Z}}

@@

Now fix an arbitrary point $x_0 \in \mathcal{X}$, and 
suppose that the kernel $K$ generates $\rho$, such that 
\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel} hold.
Then we can write \eqref{eq:within} as
\begin{equation}\label{eq:W2}
W(\{ \C_j \} )
= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg)
\end{equation}
When minimizing $W$, the first term is just a constant and does
not contribute,
therefore the problem reduces to
\begin{equation}
\label{eq:max_prob}
\max_{ \{ \C_j \} } \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
\end{equation}
Let us introduce the binary matrix $Z \in \{ 0,1 \}^{n\times k}$ 
such that 
\begin{equation}
Z_{ij} = \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise.}
\end{cases}
\end{equation}
Notice that $D = Z^T Z = \diag( n_1, n_2, \dotsc, n_k )$ contains the number
of elements in each partition. Introducing the kernel matrix
$G \in \mathbb{R}^{n\times n}$ such that
\begin{equation}
\label{eq:kernel_matrix}
G_{ij} = \kk(x_i, x_j),
\end{equation}
then \eqref{eq:max_prob} is equal to
$\max \Tr \left\{ D^{-1} Z^\top G Z \right\}$. 
Therefore, our clustering optimization problem based on energy statistics
can be formulated as
\begin{equation}\label{eq:qcqp}
\begin{aligned}
& \max_{Z} \Tr\left\{ \big( Z D^{-1/2}\big)^\top G 
\big( ZD^{-1/2} \big) 
\right\} \\
&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
$\sum_{i=1}^n Z_{ij} = n_j$, and $D = Z^\top Z$}.
\end{aligned}
\end{equation}
This is a quadratic problem with integer constraints, which is 
NP-hard. Let us now write this in terms of 
$\Zt \equiv Z D^{-1/2}$, which componentwise is 
\begin{equation}
\Zt_{ij} = \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise.}
\end{cases}
\end{equation}
We thus have
\begin{equation}
\label{eq:qcqp2}
\max_{\Zt} \Tr \left\{ \Zt^\top G \Zt \right\}  \qquad
\mbox{s.t. $\Zt \ge 0$, $\Zt^\top Z = I$, 
$\Zt \Zt^\top \bm{e} = \bm{e}$},
\end{equation}
where $\bm{e} = (1,1,\dots,1) \in \mathbb{R}^n$ is the all-ones vector,
and $G$ is the pairwise kernel matrix \eqref{eq:kernel_matrix} obtained
from \eqref{eq:kernel_semimetric}.
Therefore, to cluster data $\{ x_i \}_{i=1}^n \in \mathcal{X}$ into 
$k$ partitions,
assuming that $k$ is given, we first compute $G$  --- 
which is  defined by an arbitrary semimetric of negative 
type on $\mathcal{X}$  --- and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$.

@@

It is important to note that \eqref{eq:qcqp2} has the same
formulation as kernel $k$-means, spectral clustering, and the maximum
cut problem on graphs \cite{Dhillon}. The 
result \eqref{eq:qcqp2} brings energy statistics based clustering
into this broad picture, and \eqref{eq:qcqp2} should have 
interesting applications in
graph partitioning problems and unsupervised learning in general. 
Furthermore, and most importantly, our analysis is valid
for any space $\mathcal{X}$ equiped with a semimetric of negative type.
This method is
nonparametric since it does not assume any form of the 
distribution of the data,
contrary to $k$-means and gaussian mixture models (GMM), for example.
Also, there is no concept of cluster center involved in this approach.

@@

We can formulate an iterative algorithm to find
an approximate solution to \eqref{eq:qcqp2} on the same lines
as kernel $k$-means.
Let $t$ be the iteration time. First, precompute the kernel
matrix $G$, fix the number of clusters $k$, then
perform the following steps:
\begin{enumerate}
\item \label{initialize} 
Initialize clusters $\{ \C_1^{(0)},\dotsc,\C_k^{(0)} \}$, which
determines the label matrix $\Zt^{(0)}$.
\item \label{assignment} For each datapoint $x_i$ compute
its cluster assignment through
\begin{equation}
\label{eq:algo}
\Zt^{(t+1)}_{ij} = 
\begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $j = \argmax_\ell \tfrac{1}{n_\ell} 
\sum_{m=1}^n G_{i m} \Zt^{(t)}_{m\ell}$ } \\
0 & \mbox{otherwise.}
\end{cases}
\end{equation}
\item If converged return $\Zt^{(t+1)}$, otherwise
set $t = t+1$ and repeat step~\ref{assignment}.
\end{enumerate}
If data is $D$-dimensional, computing $G$ has complexity
$O(n^2 D)$. Step \ref{assignment} above has complexity $O(n)$ for
each point, thus total complexity $O(n^2)$. Assuming we perform $T$ 
iterations,
the total complexity of the algorithm is $O(n^2(D + T))$.
We can initialize the algorithm in step~\ref{initialize} with any method
we want, a good alternative is the initialization from $k$-means++.

@@

If data is one-dimensional and we choose
$\rho(x,y) = |x - y|$, we can actually compute 
\eqref{eq:g_def} in $O(n \log n)$ instead of $O(n^2 D)$ and find
a direct solution to \eqref{eq:minimize}. 
This is done by noticing that
\begin{equation}
|x - y|  = \mathbbm{1}(x \ge y ) (x - y) -
\mathbbm{1}(x < y) (x - y).
\end{equation}
Denote
\begin{equation}
n_j^{-} (x) = \sum_{y \in \C_j} \mathbbm{1}(y \le x), \quad  
n_j^{+} (x) = \sum_{y \in C_j} \mathbbm{1}(y > x), \quad
\mathcal{D}_j(x) = \dfrac{n_j^-(x) - n_j^+(x)}{n_j}
\end{equation}
where $n_j^+(x)$ is the number of elements in $\C_j$ which are larger
than $x \in \C_i$, and so on.
Now \eqref{eq:g_def} can be written as
\begin{equation}
\label{eq:g1d}
g(\C_i, \C_j) = 
\dfrac{1}{n_i} \sum_{x \in \C_i} \mathcal{D}_j(x) \, x - 
\dfrac{1}{n_j} \sum_{y \in \C_j} \mathcal{D}_i(y) \, y.
\end{equation}
If we merge $\C_i$ and $\C_j$ and sort the data, this expression 
can be computed in $O(n)$, where $n=n_i+n_j$.
Now for a two-class problem, we have to loop through each point
and compute \eqref{eq:within} to find the best split. Since sorting
sorting usually takes $O(n \log n)$, the complexity of
this procedure is $O(n(\log n + n)) = O(n^2)$.

>>>>

\newcommand\Energy{\mathcal{E}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
%\newcommand\Zt{\widetilde{Z}}
\newcommand\Zt{Y}

@@

Now fix an arbitrary point $x_0 \in \mathcal{X}$, and 
suppose that the kernel $K$ generates $\rho$, such that 
\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel} hold.
Then we can write \eqref{eq:within} as
\begin{equation}\label{eq:W2}
W(\{ \C_j \} )
= \dfrac{1}{2} \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y \in \C_j} \rho(x,y)
= \sum_{j=1}^k \sum_{x \in \C_j}  \bigg(
\kk(x,x) - \dfrac{1}{n_j} \sum_{y \in \C_j} \kk(x,y) \bigg)
\end{equation}
When minimizing $W$, the first term is just a constant and does
not contribute,
therefore the problem reduces to
\begin{equation}
\label{eq:max_prob}
\max_{ \{ \C_j \} } \sum_{j=1}^k \dfrac{1}{n_j} \sum_{x,y\in C_j} \kk(x,y) .
\end{equation}
Let us introduce the binary matrix $Z \in \{ 0,1 \}^{n\times k}$ 
such that 
\begin{equation}
Z_{ij} = \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise.}
\end{cases}
\end{equation}
Notice that $D = Z^T Z = \diag( n_1, n_2, \dotsc, n_k )$ contains the number
of elements in each partition. Introducing the kernel matrix
$G \in \mathbb{R}^{n\times n}$ such that
\begin{equation}
\label{eq:kernel_matrix}
G_{ij} = \kk(x_i, x_j),
\end{equation}
then \eqref{eq:max_prob} is equal to
$\max \Tr \left\{ D^{-1} Z^\top G Z \right\}$. 
Therefore, our clustering optimization problem based on energy statistics
can be formulated as
\begin{equation}\label{eq:qcqp}
\begin{aligned}
& \max_{Z} \Tr\left\{ \big( Z D^{-1/2}\big)^\top G 
\big( ZD^{-1/2} \big) 
\right\} \\
&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{j=1}^k Z_{ij} = 1$, 
$\sum_{i=1}^n Z_{ij} = n_j$, and $D = Z^\top Z$}.
\end{aligned}
\end{equation}
This is a quadratic problem with integer constraints, which is 
NP-hard. Let us now write this in terms of 
$\Zt \equiv Z D^{-1/2}$, which componentwise is 
\begin{equation}
\Zt_{ij} = \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise.}
\end{cases}
\end{equation}
We thus have
\begin{equation}
\label{eq:qcqp2}
\max_{\Zt} \Tr \left\{ \Zt^\top G \Zt \right\}  \qquad
\mbox{s.t. $\Zt \ge 0$, $\Zt^\top \Zt = I$, 
$\Zt \Zt^\top \bm{e} = \bm{e}$},
\end{equation}
where $\bm{e} = (1,1,\dots,1)^\top \in \mathbb{R}^n$ is the all-ones vector,
and $G$ is the pairwise kernel matrix \eqref{eq:kernel_matrix} obtained
from \eqref{eq:kernel_semimetric}.
Therefore, to cluster data $\{ x_i \}_{i=1}^n \in \mathcal{X}$ into 
$k$ partitions,
assuming that $k$ is given, we first compute $G$  --- 
which is  defined by an arbitrary semimetric of negative 
type on $\mathcal{X}$  --- and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$.

@@

It is important to note that \eqref{eq:qcqp2} has the same
formulation as kernel $k$-means, spectral clustering, and the maximum
cut problem on graphs \cite{Dhillon}. The 
result \eqref{eq:qcqp2} brings energy statistics based clustering
into this broad picture, and \eqref{eq:qcqp2} should have 
interesting applications in
graph partitioning problems and unsupervised learning in general. 
Furthermore, and most importantly, our analysis is valid
for any space $\mathcal{X}$ equiped with a semimetric of negative type $\rho$.
This method is
nonparametric since it does not assume any form of the 
distribution of the data,
contrary to $k$-means and gaussian mixture models (GMM), for example.
Also, there is no concept of cluster center involved in this approach.

@@

We can formulate an iterative algorithm to find
an approximate solution to \eqref{eq:qcqp2} on the same lines
as kernel $k$-means.
Let $t$ be the iteration time. First precompute the kernel
matrix $G$, fix the number of clusters $k$, then
perform the following steps:
\begin{enumerate}
\item \label{initialize} 
Initialize clusters $\{ \C_1^{(0)},\dotsc,\C_k^{(0)} \}$, which
determines the label matrix $\Zt^{(0)}$.
\item \label{assignment} For each datapoint $x_i$ compute
its cluster assignment through
\begin{equation}
\label{eq:algo}
\Zt^{(t+1)}_{ij} = 
\begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $j = \argmax_\ell \tfrac{1}{n_\ell} 
\sum_{m=1}^n G_{i m} \Zt^{(t)}_{m\ell}$ } \\
0 & \mbox{otherwise.}
\end{cases}
\end{equation}
\item If converged return $\Zt^{(t+1)}$, otherwise
set $t = t+1$ and repeat step~\ref{assignment}.
\end{enumerate}
If data is $D$-dimensional, computing $G$ has complexity
$O(n^2 D)$. Step \ref{assignment} above has complexity $O(n)$ for
each point, thus total complexity $O(n^2)$. Assuming we perform $T$ 
iterations,
the total complexity of the algorithm is $O(n^2(D + T))$.
We can initialize the algorithm in step~\ref{initialize} with any method
we want, a good alternative is the initialization from $k$-means++.

@@

If data is one-dimensional and we choose
$\rho(x,y) = |x - y|$, we can actually compute 
\eqref{eq:g_def} in $O(n \log n)$ instead of $O(n^2 D)$ and find
a direct solution to \eqref{eq:minimize}. 
This is done by noticing that
\begin{equation}
|x - y|  = \mathbbm{1}(x \ge y ) (x - y) -
\mathbbm{1}(x < y) (x - y).
\end{equation}
Denote
\begin{equation}
n_j^{-} (x) = \sum_{y \in \C_j} \mathbbm{1}(y \le x), \quad  
n_j^{+} (x) = \sum_{y \in C_j} \mathbbm{1}(y > x), \quad
\mathcal{D}_j(x) = \dfrac{n_j^-(x) - n_j^+(x)}{n_j}
\end{equation}
where $n_j^+(x)$ is the number of elements in $\C_j$ which are larger
than $x \in \C_i$, and so on.
Now \eqref{eq:g_def} can be written as
\begin{equation}
\label{eq:g1d}
g(\C_i, \C_j) = 
\dfrac{1}{n_i} \sum_{x \in \C_i} \mathcal{D}_j(x) \, x - 
\dfrac{1}{n_j} \sum_{y \in \C_j} \mathcal{D}_i(y) \, y.
\end{equation}
If we merge $\C_i$ and $\C_j$ and sort the data, this expression 
can be computed in $O(n)$, where $n=n_i+n_j$.
For a two-class problem, we have to loop through each point
and compute \eqref{eq:within} to find the best split of the data. 
Since sorting
sorting usually takes $O(n \log n)$, the complexity of
this procedure is $O(n(\log n + n)) = O(n^2)$.
